---
title: "BOO 2024 - BMD analysis"
output: html_document
date: "`r Sys.Date()`"
---

# Setup {.tabset}
```{r include=FALSE, echo=TRUE, message=FALSE}
rm(list = ls()); gc()
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80), tidy = TRUE)
```


## Load packages
### CRAN
```{r}
# Check if pacman is available and install
if(!require("pacman", quietly = T)){install.packages("pacman")}; library(pacman)

# use packman to install CRAN packages
p_load(tidyverse, ggpubr, corrr, ggfortify, ggcorrplot, ggdendro, data.table, GGally)
```



## Set directories
```{r}
# input directory
if(!dir.exists("INPUT")){
  dir.create(path = file.path(getwd(), "INPUT"))
}
input_dir <- file.path(getwd(), "INPUT")

# input directory
if(!dir.exists("OUTPUT")){
  dir.create(path = file.path(getwd(), "OUTPUT"))
}
output_dir <- file.path(getwd(), "OUTPUT")

# plot directory
if(!dir.exists("PLOT")){
  dir.create(path = file.path(getwd(), "PLOT"))
}
plot_dir <- file.path(getwd(), "PLOT")
```



# Load data {.tabset}

## Metadata
Here you should load the metadata you obtained.
* What is metadata and how many samples do you have?
- Metadata is the data that describes the countdata. There are 198 samples.
```{r}
metadata_fsample <- fread(input = file.path(input_dir, "20240516_metadata_processed.csv"))
```

## Countdata
Here you should load the raw data you obtained.
* What type of normalization was applied?
- CPM normalization.
```{r}
countdata_cpm <- fread(input = file.path(input_dir, "20240521_countdata_cpm_fsample_fprobe_sumprobe.csv"))
```


## Wrangle countdata for BMD analysis
```{r}
metadata_fsample <- metadata_fsample |>
  mutate(COMPOUND_TIME = paste0(COMPOUND,"_",TIME))

genes = countdata_cpm |> pull(GENE_SYMBOL)

compound = "CISPLATIN"

compound_time <- metadata_fsample |> 
  filter(COMPOUND == compound) |>
  pull(COMPOUND_TIME) |> 
  unique()
  

for(i in compound_time) {
  metadata_subset <- metadata_fsample |>
    filter(COMPOUND_TIME == i)
  
  countdata_subset = countdata_cpm |>
    select(GENE_SYMBOL,metadata_subset$SAMPLE_ID) |>
    column_to_rownames("GENE_SYMBOL") |>
    t() |>
    as.data.frame() |>
    rownames_to_column("SAMPLE_ID") |>
    left_join(metadata_subset |> dplyr::select(SAMPLE_ID,COMPOUND,CONCENTRATION,DOSE_LEVEL,REPLICATE)) |>
    arrange(DOSE_LEVEL,REPLICATE) |>
    dplyr::select(SAMPLE_ID, CONCENTRATION, all_of(genes)) |>
    column_to_rownames("SAMPLE_ID") |>
    t() |>
    as.data.frame() |>
    rownames_to_column("GENE_SYMBOL")
  
  fwrite(countdata_subset, paste0(input_dir, "/BMD_input_",i,"countdata_normalized_gene.txt"), row.names = F, sep = "\t")
}
```

#3 Load data in BMDExpress 2.3

Here you will load and analyse the data in BMDExpress 2.3 using the import function of a Generic geneset with a BASE2 transformation. 

First perform a William's Trend test, with p<0.05 threshold, but without any fold change threshold.
* How many genes do show a dose / concentration response for each of the timeponts?
- 4h: 199 genes, 8h: 3317 genes, 16h: 4067 genes, 24h: 8387 genes, 48h: 7115 genes, 72h: 7756 genes.
Then perform a BMD analysis with the standard settings. 
You can save the output in a text file. 
* Which five genes have the lowest BMD for each timepoint.
4h: INSIG1, KCNJ15, DNAJB9, SNAI2, WBP4.
8h: YWHAH, SEPT6, USP53, ZBTB5, RTEL1-TNFRSF6B.
16h: ACYP1, C1orf228, CFI, GOSR2, BDH1.
24h: STX12, USP5, SNAPC2, RNF5, INPP5K.
48h: ENO2, SERPINE2, ADD3, ACSL3, CAV2.
72h: FADS1, COG6, HACL1, PDE4D, AFP.

Finally perform a pathway analysis, first with REACTOME/GOterms and secondly with a predefined geneset from the RPTEC/TERT1 modules.
* Which pathways show the most significant concentration responses?
* Which modules show the most significant concentration responses?
* Are these two results pointing at similar pathways?

```{r}
install.packages("ggplot2")
library(ggplot2)
library(dplyr)
```
## Read and combine the data of functional classification
```{r}
# Read the data in R
BMD_input_modules <- read_delim(file.path(input_dir, "BMD_Input_Modules.txt"))
BMD_input_wiki <- read_delim(file.path(input_dir, "BMD_Input_Wiki.txt"), delim = "\t")
BMD_input_kegg <- read_delim(file.path(input_dir, "BMD_Input_Kegg.txt"))
BMD_input_gobp <- read_delim(file.path(input_dir, "BMD_Input_Gobp.txt"), delim = "\t")

# Combine the data
BMD_SD1 <- BMD_input_modules %>%
  select("Analysis", "BMD Standard Deviation")
BMD_SD2 <- BMD_input_wiki %>%
  select("Analysis", "BMD Standard Deviation")
BMD_SD3 <- BMD_input_kegg %>%
  select("Analysis", "BMD Standard Deviation")
BMD_SD4 <- BMD_input_gobp %>%
  select("Analysis", "BMD Standard Deviation")

BMD_SD1$Timepoint <- substring(BMD_SD1$Analysis, 11, 13)
BMD_SD2$Timepoint <- substring(BMD_SD2$Analysis, 11, 13)
BMD_SD3$Timepoint <- substring(BMD_SD3$Analysis, 11, 13)
BMD_SD4$Timepoint <- substring(BMD_SD4$Analysis, 11, 13)

BMD_SD1 <- BMD_SD1 %>%
  select("Timepoint", "BMD Standard Deviation")
BMD_SD2 <- BMD_SD2 %>%
  select("Timepoint", "BMD Standard Deviation")
BMD_SD3 <- BMD_SD3 %>%
  select("Timepoint", "BMD Standard Deviation")
BMD_SD4 <- BMD_SD4 %>%
  select("Timepoint", "BMD Standard Deviation")

BMD_SD1 <- BMD_SD1 %>%
  mutate(Pathway_database = "Modules")
BMD_SD2 <- BMD_SD2 %>%
  mutate(Pathway_database = "Wiki")
BMD_SD3 <- BMD_SD3 %>%
  mutate(Pathway_database = "KEGG")
BMD_SD4 <- BMD_SD4 %>%
  mutate(Pathway_database = "GOBP")

BMD_SD <- rbind(BMD_SD1, BMD_SD2, BMD_SD3, BMD_SD4)
BMD_list <- list(BMD_SD1, BMD_SD2, BMD_SD3, BMD_SD4)
```

## Make the boxplot with pathway databases and standard deviation
```{r}
# Load necessary libraries
library(tidyr)
library(ggplot2)

# Create the combined box plot
BMD_SD <- BMD_SD %>%
  rename(`BMD_Standard_Deviation` = `BMD Standard Deviation`)
BMD_SD$Timepoint <- factor(BMD_SD$Timepoint, levels = c("8hr", "16h", "24h", "48h", "72h"))
BMD_SD$Pathway_database <- factor(BMD_SD$Pathway_database, levels = c("Modules", "Wiki", "KEGG", "GOBP"))

ggplot(BMD_SD, aes(x = Pathway_database, y = BMD_Standard_Deviation, fill = factor(Timepoint))) + geom_boxplot() + labs(x = "Category", y = "BMD Standard Deviation") + scale_fill_discrete(name = "Timepoint") + ggtitle("Boxplot of BMD Standard Deviation by Category and Timepoint")
```
## Comparison of overlapping modules

# Modules

```{r}
#Load necessary libraries
library(dplyr)
library(ggplot2)
library(tidyr)

#Extract time point from Analysis column
BMD_input_modules <- BMD_input_modules %>%
  mutate(timepoint = str_extract(Analysis, "\\d+hr"))

#Filter and select top 10 pathways by BMD Median for each timepoint
top_pathways_M <- BMD_input_modules %>%
  filter(!is.na(timepoint)) %>%
  group_by(timepoint) %>%
  arrange(`BMD Median`) %>%
  slice(1:10) %>%
  ungroup()

#Give levels to timepoint
top_pathways_M$timepoint <- factor(top_pathways_M$timepoint, levels = c("8hr", "16hr", "24hr", "48hr", "72hr"))

#Filter between BMD SD < 10 and BMD SD > 10
under_ten_top_pathways_M <- top_pathways_M %>%
  filter(`BMD Standard Deviation` <= 10)
above_ten_top_pathways_M <- top_pathways_M %>%
  filter(`BMD Standard Deviation` > 10)

#Plot BMD SD < 10
ggplot(under_ten_top_pathways_M, aes(x = `BMD Median`, y = `GO/Pathway/Gene Set/Gene Name`, color = timepoint)) +
  geom_line() +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = `BMD Median`, xmax = `BMD Median` + `BMD Standard Deviation`), height = 0.5) +
  labs(title = "BMD Median trends of Modules over time with BMD SD < 10", 
       x = "BMD Median",
       y = "Modules",
       color = "Timepoint") +
  scale_x_log10() +
  theme_minimal()

#Plot BMD SD > 10
ggplot(above_ten_top_pathways_M, aes(x = `BMD Median`, y = `GO/Pathway/Gene Set/Gene Name`, color = timepoint)) +
  geom_line() +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = `BMD Median`, xmax = `BMD Median` + `BMD Standard Deviation`), height = 0.5) +
  labs(title = "BMD Median trends of Modules over time with BMD SD > 10", 
       x = "BMD Median",
       y = "Modules",
       color = "Timepoint") +
  scale_x_log10() +
  theme_minimal()
```
# Wiki pathways
```{r}
#Load necessary libraries
library(dplyr)
library(ggplot2)
library(tidyr)

#Extract time point from Analysis column
BMD_input_wiki <- BMD_input_wiki %>%
  mutate(timepoint = str_extract(Analysis, "\\d+hr"))

#Filter and select top 10 pathways by BMD Median for each timepoint
top_pathways_W <- BMD_input_wiki %>%
  filter(!is.na(timepoint)) %>%
  group_by(timepoint) %>%
  arrange(`BMD Median`) %>%
  slice(1:10) %>%
  ungroup()

#Give levels to timepoint
top_pathways_W$timepoint <- factor(top_pathways_W$timepoint, levels = c("8hr", "16hr", "24hr", "48hr", "72hr"))

#Filter between BMD SD < 10 and BMD SD > 10
under_ten_top_pathways_W <- top_pathways_W %>%
  filter(`BMD Standard Deviation` <= 10)
above_ten_top_pathways_W <- top_pathways_W %>%
  filter(`BMD Standard Deviation` > 10)

#Plot BMD < 10
ggplot(under_ten_top_pathways_W, aes(x = `BMD Median`, y = `GO/Pathway/Gene Set/Gene Name`, color = timepoint)) +
  geom_line() +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = `BMD Median`, xmax = `BMD Median` + `BMD Standard Deviation`), height = 0.5) +
  labs(title = "BMD Median trends of Wiki pathways over time with BMD SD < 10", 
       x = "BMD Median",
       y = "Wiki pathways",
       color = "Timepoint") +
  scale_x_log10() +
  theme_minimal()

#Plot BMD > 10
ggplot(above_ten_top_pathways_W, aes(x = `BMD Median`, y = `GO/Pathway/Gene Set/Gene Name`, color = timepoint)) +
  geom_line() +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = `BMD Median`, xmax = `BMD Median` + `BMD Standard Deviation`), height = 0.5) +
  labs(title = "BMD Median trends of Wiki pathways over time with BMD SD > 10", 
       x = "BMD Median",
       y = "Wiki pathways",
       color = "Timepoint") +
  scale_x_log10() +
  theme_minimal()

```
# KEGG
```{r}
#Load necessary libraries
library(dplyr)
library(ggplot2)
library(tidyr)

#Extract time point from Analysis column
BMD_input_kegg <- BMD_input_kegg %>%
  mutate(timepoint = str_extract(Analysis, "\\d+hr"))

#Filter and select top 10 pathways by BMD Median for each timepoint
top_pathways_K <- BMD_input_kegg %>%
  filter(!is.na(timepoint)) %>%
  group_by(timepoint) %>%
  arrange(`BMD Median`) %>%
  slice(1:10) %>%
  ungroup()

#Give levels to timepoint
top_pathways_K$timepoint <- factor(top_pathways_K$timepoint, levels = c("8hr", "16hr", "24hr", "48hr", "72hr"))

#Filter between BMD SD < 10 and BMD SD > 10
under_ten_top_pathways_K <- top_pathways_K %>%
  filter(`BMD Standard Deviation` <= 10)
above_ten_top_pathways_K <- top_pathways_K %>%
  filter(`BMD Standard Deviation` > 10)

#Plot under 10 BMD SD
ggplot(under_ten_top_pathways_K, aes(x = `BMD Median`, y = `GO/Pathway/Gene Set/Gene Name`, color = timepoint)) +
  geom_line() +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = `BMD Median`, xmax = `BMD Median` + `BMD Standard Deviation`), height = 0.5) +
  labs(title = "BMD Median trends of KEGG pathways over time with BMD SD < 10", 
       x = "BMD Median",
       y = "KEGG pathways",
       color = "Timepoint") +
  scale_x_log10() +
  theme_minimal()

#Plot above 10 BMD SD
ggplot(above_ten_top_pathways_K, aes(x = `BMD Median`, y = `GO/Pathway/Gene Set/Gene Name`, color = timepoint)) +
  geom_line() +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = `BMD Median`, xmax = `BMD Median` + `BMD Standard Deviation`), height = 0.5) +
  labs(title = "BMD Median trends of KEGG pathways over time with BMD SD > 10", 
       x = "BMD Median",
       y = "KEGG pathways",
       color = "Timepoint") +
  scale_x_log10() +
  theme_minimal()

```
# GOBP
```{r}
#Load necessary libraries
library(dplyr)
library(ggplot2)
library(tidyr)

#Extract time point from Analysis column
BMD_input_gobp <- BMD_input_gobp %>%
  mutate(timepoint = str_extract(Analysis, "\\d+hr"))

#Filter and select top 10 pathways by BMD Median for each timepoint
top_pathways_G <- BMD_input_gobp %>%
  filter(!is.na(timepoint)) %>%
  group_by(timepoint) %>%
  arrange(`BMD Median`) %>%
  slice(1:10) %>%
  ungroup()

#Give levels to timepoint and shorten one description
top_pathways_G$timepoint <- factor(top_pathways_G$timepoint, levels = c("8hr", "16hr", "24hr", "48hr", "72hr"))
top_pathways_G$`GO/Pathway/Gene Set/Gene Name` <- substring(top_pathways_G$`GO/Pathway/Gene Set/Gene Name`, 1, 80)

#Filter between BMD SD < 10 and BMD_SD > 10
under_ten_top_pathways_G <- top_pathways_G %>%
  filter(`BMD Standard Deviation` <= 10)
above_ten_top_pathways_G <- top_pathways_G %>%
  filter(`BMD Standard Deviation` > 10)

#Plot under 10 BMD SD
ggplot(under_ten_top_pathways_G, aes(x = `BMD Median`, y = `GO/Pathway/Gene Set/Gene Name`, color = timepoint)) +
  geom_line() +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = `BMD Median`, xmax = `BMD Median` + `BMD Standard Deviation`), height = 0.5) +
  labs(title = "BMD Median trends of GOBP pathways over time with BMD SD < 10", 
       x = "BMD Median",
       y = "GOBP pathways",
       color = "Timepoint") +
  scale_x_log10() +
  theme_minimal()

#Plot above 10 BMD SD
ggplot(above_ten_top_pathways_G, aes(x = `BMD Median`, y = `GO/Pathway/Gene Set/Gene Name`, color = timepoint)) +
  geom_line() +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = `BMD Median`, xmax = `BMD Median` + `BMD Standard Deviation`), height = 0.5) +
  labs(title = "BMD Median trends of GOBP pathways over time with BMD SD > 10", 
       x = "BMD Median",
       y = "GOBP pathways",
       color = "Timepoint") +
  scale_x_log10() +
  theme_minimal()

```
# Fifth percentile
```{r}
BMD_4hr <- read_delim(file.path(input_dir, "BMD_4hr.txt"), delim = "\t")
BMD_8hr <- read_delim(file.path(input_dir, "BMD_8hr.txt"), delim = "\t")
BMD_16hr <- read_delim(file.path(input_dir, "BMD_16hr.txt"), delim = "\t")
BMD_24hr <- read_delim(file.path(input_dir, "BMD_24hr.txt"), delim = "\t")
BMD_48hr <- read_delim(file.path(input_dir, "BMD_48hr.txt"), delim = "\t")
BMD_72hr <- read_delim(file.path(input_dir, "BMD_72hr.txt"), delim = "\t")

fifthp_4hr <- quantile(BMD_4hr$`Best BMD`, probs = 0.05)
fifthp_8hr <- quantile(BMD_8hr$`Best BMD`, probs = 0.05)
fifthp_16hr <- quantile(BMD_16hr$`Best BMD`, probs = 0.05)
fifthp_24hr <- quantile(BMD_24hr$`Best BMD`, probs = 0.05)
fifthp_48hr <- quantile(BMD_48hr$`Best BMD`, probs = 0.05)
fifthp_72hr <- quantile(BMD_72hr$`Best BMD`, probs = 0.05)

print(fifthp_4hr)
print(fifthp_8hr)
print(fifthp_16hr)
print(fifthp_24hr)
print(fifthp_48hr)
print(fifthp_72hr)
```
